---
title: "ClusteringDefenders"
author: "LZ"
date: "26 4 2023"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Clustering Defenders

The goal of this project is to profile defenders of the English Championship
through clustering. We will perform clustering on player stats of Season 22/23 scraped from 
FBref (<https://fbref.com/en/comps/10/stats/Championship-Stats>) with the `worldfootballR`
package (<https://github.com/JaseZiv/worldfootballR>). The data has already been 
downloaded and saved, see the Markdown document named
'Download Player Stats' for more information (also on the metrics).

We load the tibble with the player stats:
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
load(paste0(getwd(), "/data/Championship22_23_playerStats.RData"))
NL22_all
```

We filter all players that have a defending position: 

```{r}
NL22_all$Pos %>% table()
NL22_defenders <- NL22_all %>% filter(Pos %in% c("DF", "DF,FW", "DF,MF"))
dim(NL22_defenders)
```

Some players have changed teams during the season, so we sum up the total play time for them. 
```{r}

NL22_defenders <- NL22_defenders %>%  group_by(Player) %>%
  mutate(MinsPer90_total = sum(Mins_Per_90, na.rm = TRUE)) %>%
  ungroup()

```
Now we can filter those defenders with at least 180 minutes played in total during the 
season 22/23.
```{r}
NL22_defenders <- NL22_defenders %>% filter(MinsPer90_total >= 2)
```

For those who changed teams, we compute the weighted mean for each metric, where 
the weights are chosen as minutes played within teams, divided by minutes played in total.
These are the players that appear multiple times in the data, with a selection of variables.

```{r}
NL22_defenders %>% 
  filter(Player %in% NL22_defenders[duplicated(NL22_defenders$Player),  ]$Player) %>% 
  select(Player, Squad, Mins_Per_90, Tkl_Tackles, TklW_Tackles)
```
Now ´the new value for Rarmani Edmonds-Greens `Tkl_Tackles` statistics should be
\[(2.5926 \cdot 2.7 + 8.5 \cdot 1.8824)/(2.7 + 8.5).\] 
The following function does that.
```{r}
weighted_avg_duplicates <- function(df) {

  duplicated_ids <-  df[duplicated(df$Player),  ]$Player

  df_duplicates <- df %>% filter(Player %in% duplicated_ids)

  df_not_duplicated <- df %>% filter(!(Player %in% duplicated_ids))

  df_nest <- df_duplicates %>% group_by(Player, Born, Nation) %>% nest()

  weighted_avg <- function(weights, values) {
    sum(weights*values, na.rm = TRUE)/sum(weights, na.rm = TRUE)
  }
  paste_chr_info <- function(vec) {

    if(vec[1] == vec[2]) {
      return(vec[1])
    }
    else {
     paste(vec, collapse = ",")
    }
  }

  df_duplicates <- df_nest %>%
    mutate(data = purrr::map(data, ~ {

    mins.tmp <- .x$Mins_Per_90
    x_no_avg <- .x %>% select(Rk, Pos, Squad, Age)

    x_no_avg <- x_no_avg %>% summarise(across(where(is.character), paste_chr_info))
    summarise_vars <- colnames(.x)
    summarise_vars <- summarise_vars[!(summarise_vars %in% c("Rk", "Pos","Squad", "Age"))]

    xnew <- .x %>% summarise_at(summarise_vars,
                        ~ weighted_avg(weights = mins.tmp, values = .x))

    xnew <-  x_no_avg %>% bind_cols(xnew)
    xnew$Rk <- .x$Rk[1]
    xnew
    }
    )) %>%
    ungroup() %>%
    unnest(cols = data)

  df_not_duplicated %>% bind_rows(df_duplicates)

  }

```

```{r}
NL22_defenders <- weighted_avg_duplicates(NL22_defenders)

```


### Choosing relevant defending metrics

Not all available metrics are important for measuring the quality of a defender. 
We therefore define several metrics that we consider to be of central importance 
for defenders. 
First of all, we add some further metrics (Cards Per 90, Fouls per Tackle, Tackles Won).

```{r}
NL22_defenders  <- NL22_defenders %>%
  mutate(CardsPer90 = CrdY + CrdR, FoulsPerTackle = Fls/Tkl_Tackles,
         TacklesWonPercent = TklW_Tackles/Tkl_Tackles)
```

We have metrics of central importance, then some additional ones, and some measuring 
player aggressiveness:
```{r}
metrics_central <- c("Blocks_Blocks", "Sh_Blocks", "Clr", "Int",
                     "Won_Aerial Duels", "TacklesWonPercent",
                   "Tkl_percent_Challenges", "Cmp_percent_Total",
                   "Lost_Challenges", "Err", "Mins_Per_90")

metrics_add <- c("Carries_Carries", "PrgC_Carries", "Mis_Carries", 
                 "Dis_Carries", "PrgDist_Carries",
                 "CrsPA", "Cmp_percent_Long", "Cmp_percent_Short", 
                 "Cmp_percent_Medium", "PKcon", "OG")

metrics_agg <- c("Fls",  "Tkl_Tackles", "CardsPer90", "FoulsPerTackle")

metrics <- c(metrics_central, metrics_add, metrics_agg)
```

### Cluster analysis

Now we are ready for the cluster analysis. 
Our new tibble of centre backs only consists of the chosen metrics and the
Player's name and minutes played:

```{r}
cbs_ana <- NL22_defenders[ , c("Player", "Mins_Per_90", metrics)]
cbs_ana
cbs_ana <- cbs_ana[complete.cases(cbs_ana), ]
```

Since these are still many metrics, we perform a cluster analysis in order to 
reduce dimension.
```{r}
pca <- prcomp(cbs_ana[ , metrics], scale. = TRUE)
summary(pca)
```

We see that the first 10 Principal Components explain approx 80\% of variance.

We assemble a tibble with the principal component scores of the first 10 PCs for 
all players.
```{r}
rownames(pca$x) <- cbs_ana$Player
n_pcs <- 10
df_pca <- as_tibble(reshape2::melt(pca$x[, 1:n_pcs]))
df_pca <- df_pca %>% rename("Player" = Var1) %>% 
  left_join(cbs_ana %>% select(Player), by = "Player")
df_pca
```
### K-Means clustering

Since here, we have numeric metrics only and we want to find several clusters with 
similar player types, represented by an 'average' playing style, it makes sense to 
use $k$-means clustering. 
First, we look at the gap statistic and the elbow plot to find a suitable number of 
clusters.
```{r, fig.dim = c(4, 3), fig.align='center'}
set.seed(1)
factoextra::fviz_nbclust(pca$x[, 1:n_pcs], kmeans, method = "wss")
factoextra::fviz_nbclust(pca$x[, 1:n_pcs], kmeans, method = "gap_stat")
```


The gap statistic suggests that $k = 3$ is the optimal number of clusters, so we'll 
perform the clustering with 3 clusters next.

```{r}
set.seed(2)
km_clust <- kmeans(pca$x[, 1:n_pcs], centers = 3, iter.max = 50, nstart = 20)
table(km_clust$cluster)
```
 There are 80 members in cluster 1, 92 in cluster 2 and 53 in cluster 3. 
 
We can make pairwise scatterplots of the principal component (PC) scores, with the 
colour indicating cluster membership. The distributions of PC scores within the clusters
are plotted on the diagonal.

```{r, fig.dim = c(12, 10), out.width="100%", fig.align='center'}
tib_clusters <- tibble(cluster = km_clust$cluster) %>% mutate(Player = names(km_clust$cluster))
df_pca <- df_pca %>% left_join(tib_clusters)

df_pca %>% pivot_wider(names_from = Var2, values_from = value) %>%
  GGally::ggpairs(columns = 3:7, aes(colour = as.factor(cluster)), upper = "blank",
                  diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.5)), 
                  legend = c(1,1))+
  labs(fill = "Cluster")+
  theme_bw() + theme(text = element_text(size = 20), legend.position = "bottom")
```
The distributions of principal component scores are well separated between clusters 
(at least for the first two principal components). We can now try to interpret
what the clusters actually represent. 
For this, we take a look at the cluster centers and at the 
matrix of loadings.

```{r}
clust_centers <- km_clust$centers
loadings <- as_tibble(pca$rotation[ , 1:n_pcs]) %>%
  mutate_all( ~ { ifelse(abs(.x)> 0.1, round(.x,2),  "")}) %>%
  mutate(metric = rownames(pca$rotation))

knitr::kable(loadings[, c(n_pcs+1, 1:n_pcs)])
```

## Interpretation

### Principal Component Scores

* first PC
  +  high negative loads on Shots blocked, Clearances, Won Aerial Duels,
 completed pass percentage (also long + medium completed pass perc.)
  + moderately negative load on carries
  + high positive loads on lost challenges, Progressive carries, failed carries,
 dispossessed carries, Crosses into penalty area, Number of tackles
  + moderately positive load on Fouls
  + $\Rightarrow$ high negative scores on PC1 can be interpreted as having good positioning
   and defendings skills (able to clear, block shots, win aerial duels)
 while not taking too many risks (few tackles, fouls; safe passes, few failed
 or dispossessed carries)
  + confident and safe playing style
  + contarily, high scores on PC1 are related to rather bad positioning and
  + having to tackle/foul more, less safe playing style, taking more risks (also 
  for making progression)
* second PC 
  + has moderately high negative loads on Blocks in total, shots blocked, clearances,
 intercepts and aerial duels,
  + high positive loads on pass completion percentage (all distance categories),
 carries, progressive carries,
 progressive distance carried
  + balances classic defending skills and ball playing qualities,
  + high scores on PC2 related to great ball playing qualities,
  + low scores to great defending skills and less good ball playing ability.
* third PC
  + loadings suggest that PC3 measures 'intelligent'/good tackling skills or
  experience/routine 
  + positive loadings on Tkl_percent_challenges, Mins_Per_90,
 Total Tackles
  + negative loadings on conceded penalties, Fouls, CardsPer90, FoulsPerTackle
  +  having low scores on PC3 is rather unpleasant.

### Cluster Interpretation

```{r}
clust_centers[ , 1:3]
```


* first PC: separates cluster 2 and 3 quite well, cluster 1 overlaps more with 
cluster 3
  + Cluster 3 (low scores): good positioning and defending skills, safe playing style
  + Cluster 1 (balanced, skewed to low scores): also  good positioning and defending skills, 
    safe playing style
  + Cluster 2 (high scores): tendency towards more tackles and fouls, 
   more (perhaps unsuccessful) carries, likes to cross into Penalty Area 
* second PC separates cluster 1 and 3 quite well, cluster two overlaps mostly with cluster 3
  + Cluster 1 (low scores): great defending skills, fewer ball playing ability
  + Cluster 2 (moderately high scores): good ball playing/carrying abilities
  + Cluster 3 (high scores): great ball playing/carrying abilities
  
Overall, Cluster 1 is more of a 'classical' defender with great defending skills, 
good positioning and a safe playing style, without making too many progression with the 
ball or taking a lot of risk. 
Cluster 2 contains defenders that have good ball playing/carrying abilities but also 
tend to foul and tackle a lot, tendency to crosses into Penalty Area (in total, more 
willing to take a risk).
Cluster three contains defenders that have a great balance between classical defending 
skills and ball playing abilities, without taking too many risks. The have great
passing accuracy and positioning. 


Groupwise boxplots of original stats
```{r, fig.dim = c(12, 10), out.width="100%", fig.align='center'}
cbs_ana <- cbs_ana %>% select(-"Mins_Per_90") %>% left_join(tib_clusters)

cbs_ana %>% pivot_longer(cols = Blocks_Blocks:FoulsPerTackle) %>%
  ggplot(aes( x = name, y = value, fill = as.factor(cluster)))+
  geom_boxplot()+
  facet_wrap(~name, scales = "free")+
  labs(fill = "Cluster Label:", x = "")+
  theme(legend.position = "bottom", text = element_text(size = 16))
```


Finally, we get a list of players assigned to clusters.

```{r}
cluster_list <- df_pca %>% select(Player, cluster) %>% unique() %>%
  group_by(cluster) %>%
  nest()%>%
  mutate(playerlist = purrr::map_chr(data, ~ paste(unlist(.x), collapse = ", "))) %>%
  select(-data) %>% ungroup() %>% 
  arrange(cluster)

knitr::kable(cluster_list, col.names = c("Group", "Player"), align = c("l", "c"))

```




