---
title: "Clustering Defenders"
author: "LZ"
date: "26 4 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Clustering Defenders

The goal of this project is to profile defenders of the English Championship
through clustering. We will perform clustering on player stats of Season 22/23 scraped from 
FBref (<https://fbref.com/en/comps/10/stats/Championship-Stats>) with the `worldfootballR`
package (<https://github.com/JaseZiv/worldfootballR>). The data has already been 
downloaded and saved, see the Markdown document named
'Download Player Stats' for more information (also on the metrics).

We load the tibble with the player stats:
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
load(paste0(getwd(), "/data/Championship22_23_playerStats.RData"))
NL22_all
```

We filter all players that have a defending position: 

```{r}
NL22_all$Pos %>% table()
NL22_defenders <- NL22_all %>% filter(Pos %in% c("DF", "DF,FW", "DF,MF"))
dim(NL22_defenders)
```

Some players have changed teams during the season, so we sum up the total play time for them. 
```{r}

NL22_defenders <- NL22_defenders %>%  group_by(Player) %>%
  mutate(MinsPer90_total = sum(Mins_Per_90, na.rm = TRUE)) %>%
  ungroup()

```
Now we can filter those defenders with at least 180 minutes played in total during the 
season 22/23.
```{r}
NL22_defenders <- NL22_defenders %>% filter(MinsPer90_total >= 2)
```

For those who changed teams, we compute the weighted mean for each metric, where 
the weights are chosen as minutes played within teams, divided by minutes played in total.
These are the players that appear multiple times in the data, with a selection of variables.

```{r}
NL22_defenders %>% filter(Player %in% NL22_defenders[duplicated(NL22_defenders$Player),  ]$Player) %>% 
  select(Player, Squad, Mins_Per_90, Tkl_Tackles, TklW_Tackles)
```
Now ´the new value for Rarmani Edmonds-Greens `Tkl_Tackles` statistics should be
$ (2.5926 * 2.7 + 8.5* 1.8824)/(2.7 + 8.5)$. The following function does that.
```{r}
weighted_avg_duplicates <- function(df) {

  duplicated_ids <-  df[duplicated(df$Player),  ]$Player

  df_duplicates <- df %>% filter(Player %in% duplicated_ids)

  df_not_duplicated <- df %>% filter(!(Player %in% duplicated_ids))

  df_nest <- df_duplicates %>% group_by(Player, Born, Nation) %>% nest()

  weighted_avg <- function(weights, values) {
    sum(weights*values, na.rm = TRUE)/sum(weights, na.rm = TRUE)
  }
  paste_chr_info <- function(vec) {

    if(vec[1] == vec[2]) {
      return(vec[1])
    }
    else {
     paste(vec, collapse = ",")
    }
  }

  df_duplicates <- df_nest %>%
    mutate(data = purrr::map(data, ~ {

    mins.tmp <- .x$Mins_Per_90
    x_no_avg <- .x %>% select(Rk, Pos, Squad, Age)

    x_no_avg <- x_no_avg %>% summarise(across(where(is.character), paste_chr_info))
    summarise_vars <- colnames(.x)
    summarise_vars <- summarise_vars[!(summarise_vars %in% c("Rk", "Pos","Squad", "Age"))]

    xnew <- .x %>% summarise_at(summarise_vars,
                        ~ weighted_avg(weights = mins.tmp, values = .x))

    xnew <-  x_no_avg %>% bind_cols(xnew)
    xnew$Rk <- .x$Rk[1]
    xnew
    }
    )) %>%
    ungroup() %>%
    unnest(cols = data)

  df_not_duplicated %>% bind_rows(df_duplicates)

  }

```

```{r}
NL22_defenders <- weighted_avg_duplicates(NL22_defenders)

```


### Choosing relevant defending metrics

Not all available metrics are important for measuring the quality of a defender. 
We therefore define several metrics that we consider to be of central importance 
for defenders. 
First of all, we add some further metrics (Cards Per 90, Fouls per Tackle, Tackles Won).

```{r}
NL22_defenders  <- NL22_defenders %>%
  mutate(CardsPer90 = CrdY + CrdR, FoulsPerTackle = Fls/Tkl_Tackles,
         TacklesWonPercent = TklW_Tackles/Tkl_Tackles)
```

We have metrics of central importance, then some additional ones, and some measuring 
player aggressiveness:
```{r}
metrics_central <- c("Blocks_Blocks", "Sh_Blocks", "Clr", "Int", "Won_Aerial Duels", "TacklesWonPercent",
                   "Tkl_percent_Challenges", "Cmp_percent_Total","Lost_Challenges", "Err", "Mins_Per_90")

metrics_add <- c("Carries_Carries", "PrgC_Carries", "Mis_Carries", "Dis_Carries", "PrgDist_Carries",
                 "CrsPA", "Cmp_percent_Long", "Cmp_percent_Short", "Cmp_percent_Medium",
                 "PKcon", "OG")

metrics_agg <- c("Fls",  "Tkl_Tackles", "CardsPer90", "FoulsPerTackle")

metrics <- c(metrics_central, metrics_add, metrics_agg)
```

### Cluster analysis

Now we are ready for the cluster analysis. 
Our new tibble of centre backs only consists of the chosen metrics and the Player's name and minutes played:

```{r}
cbs_ana <- NL22_defenders[ , c("Player", "Mins_Per_90", metrics)]
cbs_ana
cbs_ana <- cbs_ana[complete.cases(cbs_ana), ]
```

Since these are still many metrics, we perform a cluster analysis in order to reduce dimension.
```{r}
pca <- prcomp(cbs_ana[ , metrics], scale. = TRUE)
summary(pca)
```

We see that the first 10 Principal Components explain approx 80\% of variance.

```{r}
rownames(pca$x) <- cbs_ana$Player
n_pcs <- 10
df_pca <- as_tibble(reshape2::melt(pca$x[, 1:n_pcs]))
df_pca <- df_pca %>% rename("Player" = Var1) %>% left_join(cbs_ana %>% select(Player))
df_pca
```

### hierarchical clustering

We use Ward's method (other methods yielded way worse separations).
```{r}

#dist_pcs <- dist(pca$x[ , 1:8], method = "max")

dist_pcs <- dist(pca$x[ , 1:n_pcs])

clust_pcs <- hclust(dist_pcs, method = "ward.D")
```

Now we plot the dendrogram for a first time.
```{r, fig.dim = c(10, 10)}
plot(clust_pcs, main = "Ward Linkage")
```

It seems reasonable to cut the dendrogram such that we obtain 4 clusters.
```{r, fig.dim=c(12, 8)}
cut_cl <- cutree(clust_pcs, k = 4)
clust_pcs <- as.dendrogram(clust_pcs)
dend_cl <- dendextend::color_branches(clust_pcs, k = 4)
plot(dend_cl, main = "Ward Linkage")
```

We count the number of players per cluster:

```{r}
table(cut_cl)
tib_clusters <- tibble(cluster = cut_cl) %>% mutate(Player = names(cut_cl))
```

Join the principal component scores with the obtained cluster labels, and 
plot the densities of principal component scores within clusters, as well as pair-wise
plots of the principal component scores.

```{r, fig.dim=c(10,10), out.width="80%", fig.align='center'}
df_pca <- df_pca %>% left_join(tib_clusters)

df_pca %>% pivot_wider(names_from = Var2, values_from = value) %>%
  GGally::ggpairs(columns = 3:7, aes(colour = as.factor(cluster)), upper = "blank",
                  diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.5)), legend = c(1,1))+
  labs(fill = "Cluster")+
  theme_bw() + theme(text = element_text(size = 20), legend.position = "bottom")
```

The distributions of principal component scores are well separated between clusters 
(at least for the first two principal components). We can now try to interprete 
what the scores on the differend PCs actually mean. For this, we look at the 
matrix of loadings.

```{r}
loadings <- as_tibble(pca$rotation[ , 1:n_pcs]) %>%
  mutate_all( ~ { ifelse(abs(.x)> 0.1, round(.x,2),  "")}) %>%
  mutate(metric = rownames(pca$rotation))

knitr::kable(loadings[, c(n_pcs+1, 1:n_pcs)])
```

## Interpretation

* first PC separates all clusters quite well:
 + clearest separation between clusters 1,2  and 2,3; 1,3; 3,4;
 + some overlap of clusters 1 and 4, but rather due to large variation; 
 modus is well separated; 
 + cluster 2 and 4 overlap most
* second PC separates cluster 4 from the rest.
* PC3 doesn't separate too well between the groups

* first PC
  +  high negative loads on Shots blocked, Clearances, Won Aerial Duels,
 completed pass percentage (also long + medium completed pass perc.)
  + moderately negative load on carries
  + high positive loads on lost challenges, Progressive carries, Mis_carriesm
 dispossessed carries, Crosses into penalty area, Number of tackles
  + moderately positive load on Fouls
  + $\Rightarrow$ high negative scores on PC1 can be interpreted as having good positioning
   and defendings skills (able to clear, block shots, win aerial duels)
 while not taking too many risks (few tackles, fouls; safe passes, few failed
 or dispossessed carries)
  + confident and safe playing style
  + contarily, high scores on PC1 are related to rather bad positioning and
  + having to tackle/foul more, less safe playing style
* second PC 
 + has moderately high negative loads on Blocks in total, shots blocked, clearances, intercepts and aerial duels,
 + high positive loads on pass completion percentage (all distance categories),
 carries, progressive carries,
 progressive distance carried
  + balances classic defending skills and ball playing qualities,
  + high scores on PC2 related to great ball playing qualities,
  + low scores to great defending skills and less good ball playing ability.
  + Clusters 1 and 3 are quite balanced, cluster 4 with negative scores, cluster 2
    with positive scores
* third PC
  + loadings suggest that PC3 measures 'intelligent'/good tackling skills or
  experience/routine 
  + positive loadings on Tkl_percent_challenges, Mins_Per_90,
 Total Tackles
 + negative loadings on conceded penalties, Fouls, CardsPer90, FoulsPerTackle
 +  having low scores on PC3 is rather unpleasant.


Groupwise boxplots of original stats
```{r, fig.dim = c(12, 10), out.width="80%", fig.align='center'}
cbs_ana <- cbs_ana %>% select(-"Mins_Per_90") %>% left_join(tib_clusters)

cbs_ana %>% pivot_longer(cols = Blocks_Blocks:FoulsPerTackle) %>%
  ggplot(aes( x = name, y = value, fill = as.factor(cluster)))+
  geom_boxplot()+
  facet_wrap(~name, scales = "free")+
  labs(fill = "Cluster Label:", x = "")+
  theme(legend.position = "bottom")
```


Finally, we get a list of players assigned to clusters.

```{r}
cluster_list <- df_pca %>% select(Player, cluster) %>% unique() %>%
  group_by(cluster) %>%
  nest()%>%
  mutate(playerlist = purrr::map_chr(data, ~ paste(unlist(.x), collapse = ", "))) %>%
  select(-data) %>% ungroup()

knitr::kable(cluster_list)

```
