---
title: "Download Player stats"
author: "LZ"
date: "26 4 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Download data from FBref

The goal of this project is to profile defenders of the English Championship
through clustering. We will perform clustering on player stats of Season 22/23 scraped from 
FBref (<https://fbref.com/en/comps/10/stats/Championship-Stats>) with the `worldfootballR`
package (<https://github.com/JaseZiv/worldfootballR>) in the Markdown document named
'Clustering Defenders'. 
But first, we download the 'standard', 'passing', 'defending' and 'miscellaneous' player data 
from FBref and merge them into one large tibble.

```{r, message = FALSE}
library(tidyverse)
library(worldfootballR)
```

```{r}
NL22_standard <- fb_league_stats(
  country = "ENG",
  gender = "M",
  season_end_year = 2023,
  tier = "2nd",
  non_dom_league_url = NA,
  stat_type = "standard",
  team_or_player = "player", 
  rate = purrr::rate_backoff(max_times = 5)
)
```

Here are some explanations for the measured variables: 
* `cmp` Passes completed
* `att` passes attempted
* `Cmp_percentage_*` pass completion percentage, 
 divided in Short (5-15 yards), Medium (15 - 30), Long (> 30 yards) and Total
* `totdist` total pass dist
* `prgdist` : Total distance, in yards, that completed
   passes have traveled towards the opponent's goal.
   Note: Passes away from opponent's goal are counted as zero progressive yards.
* `ast` assist
* `xAG` xG which follow a pass that assists a shot
* `xA` expected assists likelihood completed pass becomes a goal assisst given pass type,
* phase of play, location, distance
* `A-xAG` assists minus xga
* `KP` (key passes; pass leading to shot)
* `pass_final_third` passes into final third
* `PP`A passes into penalty area
* `CrsPA` crosses into penalty area
* `PrgP` progressive Completed passes that move the ball towards the opponent's
 goal line at least 10 yards from its furthest point in the last six passes,
 or any completed pass into the penalty area.
 Excludes passes from the defending 40% of the pitch

```{r}
NL22_pass <- fb_league_stats(
  country = "ENG",
  gender = "M",
  season_end_year = 2023,
  tier = "2nd",
  non_dom_league_url = NA,
  stat_type = "passing",
  team_or_player = "player",
  rate = purrr::rate_backoff(max_times = 5)
)

```

We transform all metrics that are not percentages to per 90 metrics:
```{r}
names_per_90 <- colnames(NL22_pass)[9:31]
names_per_90 <- names_per_90[ - contains("percent", vars = names_per_90)]
names_per_90

NL22_pass_per90 <- NL22_pass %>% mutate_at(names_per_90, ~ {.x/Mins_Per_90})
```

Now we download the defending stats, which have measures on the following variables:
* `Tkl` Number of players tackled
* `Tklw` Number of tackles which resulted in ball win
* `Def 3rd` Tackles in defensive 3rd
* `Mid 3rd` -- in mid 3rd
* `att 3rd` --  in attacking third
* `TkL_Challenges` Number of dribblers tackled
* `Att_Challenges` number of unsuccessful challenges + number of dribblers tackled
* `Tkl_percent_Challenges` percentage of dribblers tackled
* `Lost_Challenges` Number of unsuccessful attempts to challenge a dribbler
* `Blocks_Blocks`  Number of blocks
* `Sh_Blocks` number of shots blocked
* `Pass_Blocks` number of passes blocked
* `Int` Interceptions
* `Tkl+Int` Number of tackled plus interceptions
* `Clr` Clearances
* `Err` Errors (resulting in Shot from opponent)

```{r}
NL22_def <- fb_league_stats(
  country = "ENG",
  gender = "M",
  season_end_year = 2023,
  tier = "2nd",
  non_dom_league_url = NA,
  stat_type = "defense",
  team_or_player = "player"
)

names_per_90 <- colnames(NL22_def)[9:24]
names_per_90 <- names_per_90[ - contains("percent", vars = names_per_90)]
names_per_90

NL22_def_per90 <- NL22_def %>% mutate_at(names_per_90, ~ { .x/Mins_Per_90})
```

For possession, we have measures on

* `Touches_*` Number of touches, computed for several areas (defensive Penalty area, 
 Defensive 3rd, ...)
* `Take-On` : forcing a duel
* `Carries_Carries` Number of carries
* `TotDist_Carries` Total distance of carries in yards
* `PrgDist_Carries` Total distance in yards twrds opponents goal
* `PrgC_Carries` Number of progressive carries
* `Final_third_Carries` Carries into final third
* `CPA_Carries` carries into penalty area
* `Mis_Carries` # Failed carries
* `Dis_Carries` # Dispossessed carries
* `Rec_Receiving` # passes received (successfully)
* `PrgR_Receiving` # Progressive Passes received

```{r}

NL22_poss <- fb_league_stats(
  country = "ENG",
  gender = "M",
  season_end_year = 2023,
  tier = "2nd",
  non_dom_league_url = NA,
  stat_type = "possession",
  team_or_player = "player"
)
names_per_90 <- colnames(NL22_poss)[9:30]
names_per_90 <- names_per_90[ - contains("percent", vars = names_per_90)]
names_per_90

NL22_poss_per90 <- NL22_poss %>% mutate_at(names_per_90, ~ { .x/Mins_Per_90})
```


* `CrdY` yellow cards
* `CrdR` red cards
* `2CrdY` 2nd yellow card
* `Fls` Fouls committed
* `Fld` Fouls drawn
* `Off` Offsided
* `Crs` Crosses
* `Int` Interceptions
* `TklW` tackles won
* `PKwon` Penalty Kicks won
* `PKcon` Penalty Kicks conceded
* `OG` Own goals
* `Recov` Ball recoveries
* `Won_Aerial Duels` won aerial duels
* `Lost_Aerial Duels` lost aerial duels

```{r}
NL22_misc <- fb_league_stats(
  country = "ENG",
  gender = "M",
  season_end_year = 2023,
  tier = "2nd",
  non_dom_league_url = NA,
  stat_type = "misc",
  team_or_player = "player"
)

names_per_90 <- colnames(NL22_misc)[9:24]
names_per_90 <- names_per_90[ - contains("percent", vars = names_per_90)]
names_per_90

NL22_misc_per90 <- NL22_misc %>% mutate_at(names_per_90, ~ { .x/Mins_Per_90})

```
Now we join the tibbles (some columns are duplicated, so they are left out during the join).

```{r}
NL22_all <- NL22_standard %>% select(-c(url, Matches, CrdY, CrdR)) %>%
  left_join(NL22_def_per90 %>% select(-c(Age, url, Matches)), by = c("Rk", "Player", "Nation", "Pos", "Squad", "Born")) %>%
  left_join(NL22_pass_per90 %>%  select(-c(Age, url, Matches, Mins_Per_90, Ast)), by = c("Rk", "Player", "Nation", "Pos", "Squad", "Born")) %>%
  left_join(NL22_poss_per90 %>%  select(-c(Age, url, Matches, Mins_Per_90)), by = c("Rk", "Player", "Nation", "Pos", "Squad", "Born")) %>%
  left_join(NL22_misc_per90 %>% select(-c(Int, Age, url, Matches, Born, Mins_Per_90)), by = c("Rk", "Player", "Nation", "Pos", "Squad"))
```

We save the resulting tibble to a file named 'Championship22_23_playerStats'.
```{r}
save(NL22_all, file = paste0(getwd(), "/data/Championship22_23_playerStats.RData"))
```



